# Introduction to R {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

This book takes on a non-traditional approach to teaching R. It emphasises learning R by examples. This chapter is devoted to learning ggplot2.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
library(ggplot2)
#df
## Duration_PNES  0.097375702
## pseudostatus   0.228258071
## Hx_anxiety     0.047137943
## Hx_depression  0.005876098
## DSP            0.058153368
## seizure_burden 0.017863413
## sex            0.041313760

df<-data.frame("Predictors"=c("Duration_PNES","pseudostatus","Hx_anxiety","Hx_depression","DSP","seizure_burden","sex"),"Importance"=c(0.09737,0.22825, 0.047137,0.00487,0.058153,0.01786,0.04131))

#dimensions of data frame
dim(df)

#variables in data frame
colnames(df)

#scatter plot uses geom_point 
ggplot(df, aes(x=Predictors,y=Importance))+geom_point()

#The data frame above is unordered
#Predictors is a factor

Predictors=c("Duration_PNES","pseudostatus","Hx_anxiety","Hx_depression","DSP","seizure_burden","sex")

df<-data.frame(
  Predictors=as.factor(Predictors),
  levels=Predictors,
  "Importance"=c(0.09737,0.22825, 0.047137,0.00487,0.058153,0.01786,0.04131))

#scatter plot useing ordered factors
ggplot(df, aes(x=Predictors,y=Importance))+geom_point()

```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(df, 10), caption = 'An example table using data above',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].
