# Machine Learning Part 2

## Non-negative matrix factorisation

## Formal concept analysis

```{r fca}

library(multiplex) #Algebraic Tools for the Analysis of Multiple Social Networks
library(Rgraphviz) #plot hasse diagram

#install BiocManager::install("Rgraphviz")

edge<- read.csv("./Data-Use/Hosp_Network_geocoded.csv")
df<-edge[,c(2:dim(edge)[2])]
row.names(df)<-edge[,1] #bipartite matrix
#select columns#remove distance data
df_se<-edge[,c(2:16)]
row.names(df_se)<-edge[,1] #bipartite matrix
#south eastern hospitals
#select rows
df_se<-df_se[c(1,6,7,11,12,13,14,17,19,20,24,31,33,34,35),]
#perform Galois derivations between partially ordered subsets
#galois(df_se',labeling = "full")
gf <- galois(df_se, labeling = "reduced")
#partial ordering of concept
po<-partial.order(gf,type="galois")
diagram(po, main="Hasse diagram of partial order with reduced context") 
#lattice  diagram with reduced context
diagram.levels(po)
```

## Evolutionary Algorithm

### Simulated Annealing

### Genetic Algorithm

Genetic algorithm is a machine learning tool based on ideas from Darwin’s concept of natural selection. It is based on mutation, crossover and selection. Genetic algorithm is available in R as part of _caret_ and _GA_ libraries. Genetic algorithm can be used to optimise feature selection for regression modelling at the expense of longer running time.

## Manifold learning

### T-Stochastic Neighbourhood Embedding

T-Stochastic Neighbourhood Embedding (TSNE) is a visualisation method which seeks to transform the complex data into low (2) dimensions while maintaining the distance between neighbouring objects. This method is listed here as it is a form of data reduction method. This non-linear method is different from PCA in that the low dimensional output of TSNE are not intended for machine learning. TSNE is implemented in R as _Rtsne_.

```{r rtsne}
library(Rtsne)
```


## Deep learning

Multilayer perceptron is a type of deep learning with an inner, multiple hidden and an outer layer. It passes information in one direction from inner to hidden and outer layer and hence is referred to as feed forward artificial neural network. It trains the data using a loss function which adapt to the parameter and optimise according to the specified learning rate. Overfitting is minimised by using an L2 regularisation penalty, termed alpha. 

CNN

RNN
