# Operational Research

## Discrete Event Simulations
The example below is a based on examples provided in the _simmer_ website.

```{r simmer}
library(simmer)
library(parallel)
library(simmer.plot)
NUM_ANGIO <- 1  # Number of machines for performing ECR
ECRTIME <- 1     # hours it takes to perform ECR~ 90/60
T_INTER <- 13 # new patient every ~365*24/700 hours
SIM_TIME <- 24*30     # Simulation time in 30 days
# setup
set.seed(42)
env <- simmer()
patient <- trajectory() %>%
  log_("arrives at the ECR") %>%
  seize("removeclot", 1) %>%
  log_("enters the ECR") %>%
  timeout(ECRTIME) %>%
  set_attribute("clot_removed", function() sample(50:99, 1)) %>%
  log_(function() 
    paste0(get_attribute(env, "clot_removed"), "% of clot was removed")) %>%
  release("removeclot", 1) %>%
  log_("leaves the ECR")
env %>%
  add_resource("removeclot", NUM_ANGIO) %>%
  # feed the trajectory with 4 initial patients
  add_generator("patient_initial", patient, at(rep(0, 4))) %>%
  # new patient approx. every T_INTER minutes
  add_generator("patient", patient, function() sample((T_INTER-2):(T_INTER+2), 1)) %>%
  # start the simulation
  run(SIM_TIME)
plot(patient)
resource <- get_mon_resources(env)
plot(resource)
sum(resource$queue) #total queue size
sum(resource$queue==2) #number of people in queue
sum(resource$queue==1)
#View(get_mon_arrivals(env))
plot(env,what="arrivals",metric="flow_time")
```

## Linear Programming
There are several different libraries useful for linear programming.

```{r linear-programming}
library(lpSolveAPI)
#3 constrain 2 decision variables
x<-make.lp(3,2)
 set.column(x, 1, c(1, 1, 2)) 
  set.column(x, 2, c(3, 1, 0)) 
  set.objfn(x, c(-2, -1)) 
  set.constr.type(x, rep("<=", 3)) 
  set.rhs(x, c(4, 2, 3)) 
  solve(x)
  
```

## Forecasting
Forecasting is useful in predicting trends. In health care it can be used for estimating seasonal trends and bed requirement. Below is a forecast of mortality from COVID-19. This forecast is an example and is not meant to be used in practice as mortality from COVID also depends on the number of infected cases and the rate of infection may increase exponentially.

```{r forecast}
library(tidyverse)
library(prophet)
library(ggplot2)
library(hrbrthemes)
library(lubridate)
library(readr) #use read_csv to read csv rather than base R
covid<-read_csv("./Data-Use/Covid_Table100420.csv") 
colnames(covid)
# A data frame with columns ds & y (datetimes & metrics)
covid<-rename(covid, ds =Date, y=Total.Deaths)
covid2 <- covid[c(1:14),]
m<-prophet(covid2)#create prophet object
# Extend dataframe 12 weeks into the future
future <- make_future_dataframe(m, freq="week" , periods = 26)
# Generate forecast for next 500 days
forecast <- predict(m, future)
forecast %>% 
  select(ds, yhat, yhat_lower, yhat_upper) %>%
  tail()
# What's the forecast for July 2020?
forecasted_rides <- forecast %>%
  arrange(desc(ds)) %>%
  slice(1) %>%
  pull(yhat) %>%
  round()
forecasted_rides
# Visualize
forecast_p <- plot(m, forecast) + 
  labs(x = "", 
       y = "mortality", 
       title = "Projected COVID-19 world mortality", 
       subtitle = "based on data truncated in January 2020") +
        ylim(20000,80000)+
  theme_ipsum_rc()
forecast_p
#prophet_plot_components(m, forecast)
```
### Bed requirement

### Length of stay

### Customer churns

Customer churns or turnover is an issue of interest in marketing. The collary within healthcare are patients attendance at outpatient clinics, Insurance. The classical method used is GLM.

## Supply chains

## Health economics

### Cost

```{r hesim}
library("hesim")
library("data.table")
strategies <- data.table(strategy_id = c(1, 2))
n_patients <- 1000
patients <- data.table(patient_id = 1:n_patients,
          age = rnorm(n_patients, mean = 70, sd = 10),
          female = rbinom(n_patients, size = 1, prob = .4))
states <- data.table(state_id = c(1, 2),
                     state_name = c("Healthy", "Sick")) 
# Non-death health states
tmat <- rbind(c(NA, 1, 2),
              c(3, NA, 4),
              c(NA, NA, NA))
colnames(tmat) <- rownames(tmat) <- c("Healthy", "Sick", "Dead")
transitions <- create_trans_dt(tmat)
transitions[, trans := factor(transition_id)]
hesim_dat <- hesim_data(strategies = strategies,
                        patients = patients, 
                        states = states,
                        transitions = transitions)
print(hesim_dat)
```

Data from WHO on mortality rate can be extracted directly from WHO or by calling _get_who_mr_ in _heemod_ library.

```{r heemod}
library(heemod)

```

There are several data in BCEA library: Vaccine.[@Baio2012]

```{r bcea}
library(BCEA)
#help (package="BCEA")
data(Vaccine,package = "BCEA")
View(vaccine)
ints=c("Standard care","Vaccination")
m <- bcea(e,c,ref=2,interventions=ints)
summary(m)
```