# Operational Research

## Queueing theory

Queueing theory describes the movement of a queue such as customer arrival in bank, shop or emergency department. It seeks to balance supply and demand for a service. It begun with the study queue waiting on Danish telephones in 1909.

Little's theorem describes the linear relationship between the number of customer _L_ to the customer arrival rate $\lambda$ and the customer served per time peiod, $\mu$. This can also be used to determine the number of beds needed for coronary care unit given 4 patients being admitted to cardiology unit, one of whom needs to be admitted to coronary care unit and would stay for an average of 3 days.

Queueing system is described in terms of Kendall’s notation, M/M/c/k, with exponential arrival time. Using this terminology, MM1 system has 1 server and infinite queue. An MM/2/3 system has 2 (c) servers and 1 (k-c) position in the queue. The M refers to Markov chain. 

An example of a single server providing full service is a car wash. Example of a single multiphase server include different single stations in bank of withdrawing, deposit, information  A counter at the airport or train station for economy and business passengers is considered multiserver single phase queue. A laundromat with different queues for washing and drying is an example of multiphase multiservers.

A traditional queue at a shop can also be seen as first in first out with the first customer served first and leave first. An issue with FIFO is that people may queue early such as overnight queue for the latest iPhone. Alternatives include last in first out queue and priority queueing in emergency department.


Let's create a simple queue with 2 customers arriving per minute and 3 customers served per minute. The PO is the probability that the server is idle. 

```{r 08-operational-research-1, warning=F}

library(queueing)

lambda <- 2 # 2 customers arriving per minute
mu <- 3 # 3 customers served per minute

# MM1 

mm1 <- NewInput.MM1(lambda = 2, mu = 3, n = 0)

# Create queue class object
mm1_out <- QueueingModel(mm1)

# Report
Report(mm1_out)

# Summary
summary(mm1_out)

```
```{r 08-operational-research-1-1, warning=F}

curve(dpois(x, mm1$lambda),
      from = 0, 
      to = 20, 
      type = "b", 
      lwd = 2,
      xlab = "Number of customers",
      ylab = "Probability",
      main = "Poisson Distribution for Arrival Process",
      ylim = c(0, 0.4),
      n = 21)

```

Lets examine M/M/3 queue with exponential inter-arrival times, exponential service times and 3 servers.

```{r 08-operational-research-2, warning=F}

library(queuecomputer)

n <- 100
arrivals <- cumsum(rexp(n, 1.9))
service <- rexp(n)

mm3 <- queue_step(arrivals = arrivals, service = service, servers = 3)

```

Plot the arrival and departure times

```{r 08-operational-research-2-1, warning=F}
plot(mm3)[1]

```
Plot waiting time

```{r 08-operational-research-2-2, warning=F}
plot(mm3)[2]

```

Plot customer in queue

```{r 08-operational-research-2-3, warning=F}
plot(mm3)[3]

```

Plot customer and server status

```{r 08-operational-research-2-4, warning=F}
plot(mm3)[4]

```

Plot arrival and departure time

```{r 08-operational-research-2-5, warning=F}
plot(mm3)[5]

```
## Discrete Event Simulations

Discrete event simulation can be consider as modeling a complexity of system with multiple processes over time. This is different from continuous modeling of a system which evolve continuously with time. Discrete event simulation can be apply to the study of queue such as bank teller with a first in first out system.

### Simulate capacity of system

The example below is a based on examples provided in the _simmer_ website for laundromat.

```{r 08-operational-research-3, warning=F}
library(simmer)
library(parallel)
library(simmer.plot)

NUM_ANGIO <- 1  # Number of machines for performing ECR
ECRTIME <- 1     # hours it takes to perform ECR~ 90/60
T_INTER <- 13 # new patient every ~365*24/700 hours
SIM_TIME <- 24*30     # Simulation time over 30 days

# setup
set.seed(42)
env <- simmer()
patient <- trajectory() %>%
  log_("arrives at the ECR") %>%
  seize("removeclot", 1) %>%
  log_("enters the ECR") %>%
  timeout(ECRTIME) %>%
  set_attribute("clot_removed", function() sample(50:99, 1)) %>%
  log_(function() 
    paste0(get_attribute(env, "clot_removed"), "% of clot was removed")) %>%
  release("removeclot", 1) %>%
  log_("leaves the ECR")
env %>%
  add_resource("removeclot", NUM_ANGIO) %>%
  # feed the trajectory with 4 initial patients
  add_generator("patient_initial", patient, at(rep(0, 4))) %>%
  # new patient approx. every T_INTER minutes
  add_generator("patient", patient, function() sample((T_INTER-2):(T_INTER+2), 1)) %>%
  # start the simulation
  run(SIM_TIME)
```

Plot the schematics of the simulation.

```{r 08-operational-research-3-1, warning=F}
plot(patient)
```
Plot resource usage

```{r 08-operational-research-3-2, warning=F}
resource <- get_mon_resources(env)
plot(resource)
```
Total queue size

```{r 08-operational-research-3-3, warning=F}
sum(resource$queue) #total queue size
```
Number of people in queue of size 1

```{r 08-operational-research-3-4, warning=F}
sum(resource$queue==1) #number of people in queue
```

Number of peope in queue of size 2

```{r 08-operational-research-3-5}
sum(resource$queue==2)
```

Plot arrival versus flow

```{r 08-operational-research-3-6, warning=F}
#View(get_mon_arrivals(env))
plot(env,what="arrivals",metric="flow_time")
```

Next we simulate the process of running a stroke code. In this simulation the activities occur sequentially.

```{r 08-operational-research-4}
#simulate ST6
patient <- trajectory("patients' path") %>%
  ## add an intake activity - nurse arrive first on scene in ED to triage
  #seize specify priority
  seize("ed nurse", 1) %>%
  #timeout return one random value from 5 mean+/-5 SD 
  timeout(function() rnorm(1,5,5)) %>%
  release("ed nurse", 1) %>%
  
  ## add a registrar activity - stroke registrar arrive after stroke code
  seize("stroke reg", 1) %>%
  timeout(function() rnorm(1, 10,5)) %>%
  release("stroke reg", 1) %>%
  
  #add CT scanning - the process takes 15 minutes
  seize("CT scan", 1) %>%
  timeout(function() rnorm(1, 15,15)) %>%
  release("CT scan", 1) %>%
  
    #stroke reg re-enter
    #seize("stroke reg", 1) %>%
    #timeout(function() rnorm(1, 10,5)) %>%
    #release("stroke reg", 1) %>%
  #branch
  
  ## add stroke consultant - to review scan and makes decision
  
  seize("stroke consultant", 1) %>%
  timeout(function() rnorm(1, 5,10)) %>%
  release("stroke consultant", 1) %>%
  
  ## add a thrombectomy decision activity
  seize("inr", 1) %>%
  timeout(function() rnorm(1, 5,5)) %>%
  release("inr", 1) 

envs <- mclapply(1:100, function(i) {
  simmer("SuperDuperSim") %>%
    add_resource("ed nurse", 1) %>%
    add_resource("stroke reg", 1) %>%
    add_resource("CT scan", 1) %>%
    add_resource("stroke consultant", 1) %>%
    add_resource("inr", 1) %>%
    add_generator("patient", patient, function() rnorm(1, 10, 2)) %>%
    run(100) %>%
    wrap()
})
```

Plot patient flow

```{r 08-operational-research-4-1, warning=F}

plot(patient)

```

```{r 08-operational-research-4-2, warning=F}
#plot.simmer
resources <- get_mon_resources(envs)

#
p1<-plot(resources, metric = "usage", 
         c("ed nurse","stroke reg","CT scan", "stroke                         consultant","inr"), 
         items = "serve")

#resource usage
p2<-plot(get_mon_resources(envs[[6]]), metric = "usage", "stroke consultant", items = "server", steps = TRUE)

#resource utilisation
p3<-plot(resources, metric="utilization", c("ed nurse", "stroke reg","CT scan"))

#Flow time evolution
arrivals <- get_mon_arrivals(envs)
p4<-plot(arrivals, metric = "flow_time")

#combine plot
gridExtra::grid.arrange(p1,p2, p3,p4)
```


```{r 08-operational-research-4-3, warning=F}
#
activity<-get_mon_arrivals(envs)

plot(activity,metric="activity_time")
```

Now we will fork another path in the patient flow

```{r 08-operational-research-4-4, warning=F}

```


### Queuing network

```{r 08-operational-research-5}

mean_pkt_size <- 100        # bytes
lambda1 <- 2                # pkts/s
lambda3 <- 0.5              # pkts/s
lambda4 <- 0.6              # pkts/s
rate <- 2.2 * mean_pkt_size # bytes/s

# set an exponential message size of mean mean_pkt_size
set_msg_size <- function(.)
  set_attribute(., "size", function() rexp(1, 1/mean_pkt_size))

# seize an M/D/1 queue by id; the timeout is function of the message size
md1 <- function(., id)
  seize(., paste0("md1_", id), 1) %>%
  timeout(function() get_attribute(env, "size") / rate) %>%
  release(paste0("md1_", id), 1)

```


```{r 08-operational-research-5-1}

to_queue_1 <- trajectory() %>%
  set_msg_size() %>%
  md1(1) %>%
  leave(0.25) %>%
  md1(2) %>%
  branch(
    function() (runif(1) > 0.65) + 1, continue=c(F, F),
    trajectory() %>% md1(3),
    trajectory() %>% md1(4)
  )

to_queue_3 <- trajectory() %>%
  set_msg_size() %>%
  md1(3)

to_queue_4 <- trajectory() %>%
  set_msg_size() %>%
  md1(4)

```

```{r 08-operational-research-5-3}
env <- simmer()
for (i in 1:4) env %>% 
  add_resource(paste0("md1_", i))
env %>%
  add_generator("arrival1_", to_queue_1, function() rexp(1, lambda1), mon=2) %>%
  add_generator("arrival3_", to_queue_3, function() rexp(1, lambda3), mon=2) %>%
  add_generator("arrival4_", to_queue_4, function() rexp(1, lambda4), mon=2) %>%
  run(4000)

```

```{r 08-operational-research-5-4}
res <- get_mon_arrivals(env, per_resource = TRUE) %>%
  subset(resource %in% c("md1_3", "md1_4"), select=c("name", "resource"))

arr <- get_mon_arrivals(env) %>%
  transform(waiting_time = end_time - (start_time + activity_time)) %>%
  transform(generator = regmatches(name, regexpr("arrival[[:digit:]]", name))) %>%
  merge(res)

aggregate(waiting_time ~ generator + resource, arr, function(x) sum(x)/length(x))
get_n_generated(env, "arrival1_") + get_n_generated(env, "arrival4_")
aggregate(waiting_time ~ generator + resource, arr, length)

```

## Linear Programming

Linear programming is an optimisation process to maximise profit and minimise cost with multiple parts of the model having linear relationship. There are several different libraries useful for linear programming. The _lpSolve_ library is used here as illustration.

```{r 08-operational-research-6}
library(lpSolve)

#solve using linear programming
n <-2.5 # Numbers of techs. 1 EFT means a person is employed for 40 hours a week and 0.5 EFT means a person is employed for 20 hours a week.
set_up_eeg <- 40 # 40 minutes
to_do_eeg <- 30 # 30 minutes
clean_equipment <- 20 # 20 minutes
annotate_eeg <- 10 # 10 minutes

# put some error for EEG time
# if error=1 that mean NO errors happen
error <- 0.8 #change from 0.93

#Calculate time for EEG in hour
eeg_case_time <- ((set_up_eeg+to_do_eeg+clean_equipment+annotate_eeg)/60)*error

# limit for EEG per day
# we can put different limits for EEGs
limit_eeg <- round(8*eeg_case_time, digits = 0)

#s[i] - numbers of cases for each i-EEG's machines 
#Setting the coefficients of s[i]-decision variables
#In a future can put some efficiency or some cost
objective.in=c(1,1,1,1,1)

#Constraint Matrix
const.mat=matrix(c(1,0,0,0,0,
                   0,1,0,0,0,
                   0,0,1,0,0,
                   0,0,0,1,0,
                   0,0,0,0,1,
                   1,1,1,1,1),nrow = 6,byrow = T)

#defining constraints
const_num_1=limit_eeg  #in cases
const_num_2=limit_eeg  #in cases
const_num_3=limit_eeg  #in cases
const_num_4=limit_eeg  #in cases
const_num_5=limit_eeg  #in cases
const_res= n*7 # limit per sessions

#RHS for constraints
const.rhs=c(const_num_1,const_num_2,const_num_3,const_num_4,const_num_5, const_res)

#Direction for constraints
constr.dir <- rep("<=",6)

#Finding the optimum solution
opt=lp(direction = "max",objective.in,const.mat,constr.dir,const.rhs)
#summary(opt)

#Objective values of s[i]

opt$solution 
  
  
```

Estimate for day (Value of objective function at optimal point)

```{r 08-operational-research-6-1, echo=FALSE, message=FALSE}
opt$objval
```

Estimate EEG per month based on staff EFT- only 2.5 

```{r 08-operational-research-6-2, echo=FALSE, message=FALSE}

#Urgent cases estimate
saturday <-2
sunday <- 2
estimate_week <- 5*opt$objval + saturday + sunday
estimate_month <- 4*estimate_week
estimate_month
```
Assuming that the time spend on a report by neurologists (1 report = 30 min) then in a 3.5 hour session a neurologist can report 7 EEG.


```{r 08-operational-research-6-3}

neurologist_session=estimate_week/7

neurologist_session


```

## Forecasting

Forecasting is useful in predicting trends. In health care it can be used for estimating seasonal trends and bed requirement. Below is a forecast of mortality from COVID-19 in 2020. This forecast is an example and is not meant to be used in practice as mortality from COVID depends on the number of factors including infected cases, age, socioeconomic group, and comorbidity. 


```{r 08-operational-research-7}
library(tidyverse)
library(prophet)
library(hrbrthemes)
library(lubridate)
library(readr) #use read_csv to read csv rather than base R

covid<-read_csv("./Data-Use/Covid_Table100420.csv") 
colnames(covid)

# A data frame with columns ds & y (datetimes & metrics)
covid<-rename(covid, ds =Date, y=Total.Deaths)
covid2 <- covid[c(1:12),]

m<-prophet(covid2)#create prophet object
# Extend dataframe 12 weeks into the future
future <- make_future_dataframe(m, freq="week" , periods = 26)
# Generate forecast for next 500 days
forecast <- predict(m, future)

# What's the forecast for July 2020?
forecasted_rides <- forecast %>%
  arrange(desc(ds)) %>%
  dplyr::slice(1) %>%
  pull(yhat) %>%
  round()
forecasted_rides

# Visualize
forecast_p <- plot(m, forecast) + 
  labs(x = "", 
       y = "mortality", 
       title = "Projected COVID-19 world mortality", 
       subtitle = "based on data truncated in January 2020") +
        ylim(20000,80000)+
  theme_ipsum_rc()
#forecast_p

```

### Bed requirement

### Length of stay

### Customer churns

Customer churns or turnover is an issue of interest in marketing. The corollary within healthcare is patients attendance at outpatient clinics, Insurance. The classical method used is GLM.


## Process mapping

```{r 08-operational-research-8}
library(DiagrammeR)
a.plot<-mermaid("
        graph TB
        
        A((Triage))
        A-->|2.3 hr|B(Imaging-No Stroke Code)
        A-->|0.6 hr|B1(Imaging-Stroke Code)
        
        B-->|14.6 hr|B2(Dysphagia Screen)
        B1-->|no TPA 10.7 hr|B2(Dysphagia Screen)

        C(Stop NBM)
        B2-->|0 hr|C
      
        C-->|Oral route 1.7 hr|E{Antithrombotics}
        
        D1-->|7.5 hr|E
        B-->|PR route 6.8 hr|E
        B1-->|PR route 3.8 hr|E
    
        B1-->|TPA 24.7 hr|D1(Post TPA Scan)
    
        style A fill:#ADF, stroke:#333, stroke-width:2px
        style B fill:#9AA, stroke:#333, stroke-width:2px
        style B2 fill:#9AA, stroke:#333, stroke-width:2px
        style B1 fill:#879, stroke:#333, stroke-width:2px
        style C fill:#9AA, stroke:#333, stroke-width:2px
        style D1 fill:#879, stroke:#333, stroke-width:2px
        style E fill:#9C2, stroke:#9C2, stroke-width:2px
        ") 
a.plot
```

```{r 08-operational-research-9}
library(bupaR)
```

## Supply chains

## Health economics

### Cost

```{r 08-operational-research-10}

library("hesim")
library("data.table")

strategies <- data.table(strategy_id = c(1, 2))
n_patients <- 1000
patients <- data.table(patient_id = 1:n_patients,
          age = rnorm(n_patients, mean = 70, sd = 10),
          female = rbinom(n_patients, size = 1, prob = .4))
states <- data.table(state_id = c(1, 2),
                     state_name = c("Healthy", "Sick")) 
# Non-death health states
tmat <- rbind(c(NA, 1, 2),
              c(3, NA, 4),
              c(NA, NA, NA))
colnames(tmat) <- rownames(tmat) <- c("Healthy", "Sick", "Dead")
transitions <- create_trans_dt(tmat)
transitions[, trans := factor(transition_id)]
hesim_dat <- hesim_data(strategies = strategies,
                        patients = patients, 
                        states = states,
                        transitions = transitions)
print(hesim_dat)
```

Data from WHO on mortality rate can be extracted directly from WHO or by calling _get_who_mr_ in _heemod_ library.

```{r 08-operational-research-11}

library(heemod)

```

There are several data in BCEA library such as Vaccine.

```{r 08-operational-research-12}
library(BCEA)

#use Vaccine data from BCEA
data(Vaccine)

ints=c("Standard care","Vaccination")

# Runs the health economic evaluation using BCEA
m <- bcea(
      e=eff,
      c=cost,               # defines the variables of 
                            #  effectiveness and cost
      ref=2,                # selects the 2nd row of (e, c) 
                            #  as containing the reference intervention
      interventions=treats, # defines the labels to be associated 
                            #  with each intervention
      Kmax=50000,           # maximum value possible for the willingness 
                            #  to pay threshold; implies that k is chosen 
                            #  in a grid from the interval (0, Kmax)
      plot=TRUE             # plots the results
)

```