# Graph Theory
Graph based methods are used daily without realisation of their origin. Examples include Google, Facebook social network, and roads. These methods have emerged as tools for interpreting and analysing connected network structures [@c38a1e53f6a2457198c204405a3d7e00]. For example,  the brain structures and their association with disability can be framed as a network analysis.

Matrix can be used to represent graphical data, relevant for the network analysis. An adjacency matrix represents the adjacency between different nodes. The nodes are represented in rows and colums. A value of one represent that node A is adjancence to node B. This graph is symmetrical as the value is also assigned between B and A. The diagonal of this graph consists of zeros as the relationship between A and A is zero. Direction can be introduced to a graph by representing the direction from A to B as one. In this case, a value of zero is attached from B to A. 

A graph consists of vertices (nodes) and edges (links). The edges can have direction in which case it is known as directed graph (digraph). Edge direction indicates flow from a source node to a destination node. The nodes and their directed edges can be represented as an adjacency matrix. A directed graph has an asymmetric adjacency matrix.


## Special graphs 

### Laplacian matrix
The Laplacian matrix is the difference between the degree and adjacency matrix. The degree matrix represents the number of direct connection of a node. The diagonal of Laplacian matrix retains the diagonal values of the degree matrix (since the diagonal of the adjacency matrix consists of zeroes). The smallest eigenvalue of the Laplacian describes whether the graph is connected or not. The second eigenvalue of the Laplacian matrix is the algebraic connectivity or Fiedler value. High Fiedler value indicates greater number of connected components and consequently, resilience to breakdown in flow of information between members[@conf/networking/JamakovicM08] .

### Bimodal (bipartite) graph
Bimodal graphs are of interest in social network and in analysis of food ecosystem in nature. A well studied example is the _Zachary_ karate club network.This dataset is included in the _igraphdata_ package.
```{r bimodal}
library(bipartite)
edge<- read.csv("./Data-Use/Hosp_Network_geocoded.csv")
#community structure of a network as distinct clusters of interactions
df<-edge[,c(2:dim(edge)[2])]
row.names(df)<-edge[,1] 
#select columns to remove distance data
df_se<-edge[,c(2:16)]
row.names(df_se)<-edge[,1] 
#select rows to subset south eastern hospitals
df_se<-df_se[c(1,6,7,11,12,13,14,17,19,20,24,31,33,34,35),]
Nedge3<-computeModules(df_se)
plotModuleWeb(Nedge3) #module web object
```
```{r bipartite}
library(latentnet)
data(davis)
davis.fit<-ergmm(davis~bilinear(d=2)+rsociality)
plot(davis.fit,pie=TRUE,rand.eff="sociality",labels=TRUE)
#davis[,1:10]
```

## Centrality Measures
The nodes in this case are variables such as ASPECTS regions, demographic and risk factors. This graph below uses data from a paper on the use of PageRank. In Figure 1 of that papera graph of 4 variables were used to illustrate the PageRank method for searching brain regions related to disability [@pmid25961856].
```{r eval=FALSE}
library(igraph)
aspects<-graph(c("M1","M2","M2","Disability","tpa","M1","M2","M1","tpa","M2"))
plot(aspects) #aspects is the name of the graph
#degree
d<-degree(aspects)
#closeness
cl<-closeness(aspects)
#betweenness
b<-betweenness(aspects)
#page rank
p<-page.rank(aspects)
df<-data.frame("degree"=round(d,2),"closeness"=round(cl,2),"betweenness" =round(b,2),"PageRank" =round(p$vector,2))
knitr::kable(df)
```
## Local centrality measures
Centrality measures assign a measure of “importance” to nodes and can therefore indicate whether some nodes are more critical than others in a given network. When network nodes represent variables, centrality measures may indicate relevance of variables to a model.

The simplest centrality measure, degree centrality, is the count of links for each node and is a purely local measure of importance. Node strength, used in weighted networks, is the sum of weights of edges entering or leaving (or both) the node. Other measures, such as betweenness centrality, describe more global structure - the degree of participation of a node as conduit of information between other nodes in a network.

## Global centrality measures

### Page Rank
PageRank is one member of a family of graph eigenvector centrality measures, all of which incorporate the idea that the score of a node depends, at least in part, on the scores of neighbors connecting to the node. Thus a page may have a high PageRank score if many pages link to it, or if a few important or authoritative pages link to it. 

Others include eigenvector centrality (which works best with undirected graphs), alpha centrality and Katz centrality. PageRank uses a different scaling for connections (by the number of links leaving the node) and importance is based on incoming connections rather than outgoing connections[@brin98anatomy] . Eigenvector centrality measure a node’s centrality in terms of node parameters and centrality of neighboring nodes. 

PageRank has several differences with respect to other eigenvector centrality methods, expanded below, which make it better suited for digraphs. PageRank was originally described in terms of a web user/surfer randomly clicking links, and the PageRank of a web page corresponds to the probability of the random surfer arriving at the page of interest. The model of the random surfer used in the PageRank computation includes a damping factor, which represents the chance of the random surfer becoming bored and selecting a completely different page at random (teleporting to a random page). Similarly, if a page is a sink (i.e. has no outgoing links), then the random web surfer may click on to a random page.

A number of different approaches are available for computing the PageRank for nodes in a network. The conceptually simplest is to assign an equal initial score to each node, and then iteratively update PageRank scores. This is easy to perform algebraically for a small number of nodes but can take a long time with larger data. In practice, it is performed using eigenvector methode. PageRank analysis can be performed using _igraph_ package. [@pmid25961856]


Graph based methods have emerged as tools for interpreting and analysing connected network structures and in this case network structures associated with disability.These types of analysis are attractive because they assess the connectedness of each region of interest (ROI) with respect to other ROIs over the entire brain network.Eigenvector centrality methods have been used to explore connectedness of brain regions. PageRank is a variant of eigenvector centrality and is an ideal method for analysis ofdirected graph (the edges between adjacent nodes(regions) have direction). This method was initially developed as the basis of the search engine for Google. PageRank offered a considerable improvement over pure text based methods in ranking search results,and had the advantage of being content independent (i.e. the search is based on links between the web pages).  PageRank emphasises web pages based on the number of links directed to a page and the importance of the sources of those links. Thus a small number of links from influential pages can greatly enhance the importance of the destination page. 


## Community

## Visualising graph
There are many packages for visualising graph such as _igraph_, _sna_, _ggraph_, _Rgraphviz, _visnetwork_, _networkD3_.

### Visnetwork
```{r visnetwork}
library(igraph)
library(RColorBrewer)
library(visNetwork)
edge<- read.csv("./Data-Use/TPA_edge.csv") #3 columns:V1 V2 time 
node<-read.csv("./Data-Use/TPA_node.csv")
df<-graph_from_data_frame(d=edge[,c(1,2)],vertices = node,directed=F)
V(df)$type<-V(df)$name %in% edge[,1]
#assign color
vertex.label<-V(df)$membership
# Generate colors based on type:
colrs <- c("gray50", "gold")
V(df)$color <- colrs[V(df)$membership]
#assign shape
shape <- c("circle", "square")
V(df)$shape<-shape[V(df)$membership] 
#extract data for visNetwork
data<-toVisNetworkData(df)
visNetwork(nodes=data$nodes,edges=data$edges, main="TPA ECR Network")%>% visNodes(color = list(hover = "red")) %>% visInteraction(hover = TRUE)
```

### Large graph
There are very few softwares capable of handling very large graph with the exception of _Gephi_ , _Cytoscape_ and _Neo4J_.

## Social Media and Network Analysis
Social media platform such as twitter, youtube, facebook and instagram are rich source of information for graph theory analysis as well as textmining. The following section only covers twitter and youtube as both are accessible to the public. There's restricted access for facebook and instagram.

### Twitter

### Youtube
