# Geospatial analysis

This section deals with the use of geospatial analysis in healthcare. There is a detailed tutorial online available [here](./https://richardbeare.github.io/GeospatialStroke/
)

There are several packages avalable for obtaining geocode. The _tmaptools_ package provide free geocoding using OSM overpass API. Both _ggmap_ and _googleway_ access Google Maps API and will require a key and payment for access.

## Geocoding

```{r eval=FALSE}
library(tmaptools)
mmc<-geocode_OSM ("monash medical centre, clayton")
mmc
```

The equivalent code in _ggmap_ is provided below.

```{r eval=FALSE}
library(ggmap)
register_google(key="Your Key")
#geocode
geocode ("monash medical centre, clayton")
#trip
mapdist("5 stud rd dandenong","monash medical centre")
```

The next demonstration is the extraction of geocodes from multiple addresses embedded in a column of data wihin a dataframe. This is more efficient compared to performing geocoding line by line. An example is provided on how to create your own icon on the _leaflet_ document.

```{r geocode}
library(dplyr)
library(tidyr) #unite verb from tidyr
library(readr)
library(tmaptools)
library(leaflet)
clinic<-read_csv("./Data-Use/TIA_clinics.csv")
clinic2<-clinic %>%  
  unite ("address",City:Country,sep = ",")%>%   filter(!is.na(`Clinic-Status`)) #remove NA cells
#clinics_geo<-geocode_OSM(clinic2$address) #geocode TIA clinic
#save(clinics_geo,file="./Data-Use/TIAclinics_geo.Rda") #save data
#merge 2 files with different column name
load("./Data-Use/TIAclinics_geo.Rda")
clinics_geo<-left_join(clinics_geo,clinic2, by=c("query"="address"))
#create icon markers
#icon markers
icons_blue <- awesomeIcons(
  icon= 'medkit',
  iconColor = 'black',
  library = 'ion',
  markerColor = "blue"
)
icons_red <- awesomeIcons(
  icon= 'medkit',
  iconColor = 'black',
  library = 'ion',
  markerColor = "red"
)
#subset 
clinics_geo_active<-clinics_geo %>%filter(`Clinic-Status`=="Active")
clinics_geo_inactive<-clinics_geo %>%filter(`Clinic-Status` !="Active")
m<-leaflet(data=clinics_geo) %>% 
  addTiles() %>% #default is OSM
    addAwesomeMarkers(lat=clinics_geo_active$lat,lng=clinics_geo_active$lon,icon=icons_blue,label = ~as.character(clinics_geo_active$query) ) %>%
  addAwesomeMarkers(lat=clinics_geo_inactive$lat,lng=clinics_geo_inactive$lon,icon=icons_red,label = ~as.character(clinics_geo_inactive$query) ) 

#make pics using mapshot
mapview::mapshot(m, url = paste0(getwd(),file="./Data-Use/TIAclinic_world.html"), file = paste0(getwd(), "./Data-Use/TIAclinic_world.png"))
m
```

Googleway had the flexibility of easily interrogating Google Maps API for time of trip and traffic condition.

```{r eval=FALSE}
library(googleway)
library(googleway)
key="Your Key"
#trip to MMC
#traffic model can be optimistic, best guess, pessimistic
google_distance("5 stud rd dandenong","monash medical centre", key=key, departure_time=as.POSIXct("2019-12-03 08:15:00 AEST"),traffic_model = "optimistic")
```

There are several methods for reading the shapefile data. Previously _rgdal_ library was used. This approach creates files which can be described as S4 object in that there are slots for different data. The simple feature _sf_ approach is much easier to handle and the data can easily be subsetted and merged if needed.

## Thematic map
In the first chapter we provided a thematic map example with _ggplot2_. here we will illustrate with _mapview_.

```{r map}
library(mapview)
library(sf)
Aust<-st_read("./Data-Use/GCCSA_2016_AUST.shp") 
mapview(Aust, zcol="AREASQKM16")
```

## Spatial regression

## Spatio-temporal regression
