--- 
title: "Applications of R in Healthcare"
author: "Thanh Phan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: GNtem2/HealthcareRbook
description: "This is a book describing the use of R in Healthcare. It is aimed at beginners of R language. The majority of the examples are taken from works in Neurology. Where possible data from other disease such as heart disease, cancer and vaccine are used. The ideas and principles can be applied to other aspects of Healthcare. The output format for this example is bookdown::gitbook."
---

# Preamble

Placeholder



<!--chapter:end:index.Rmd-->


# Introduction to R {#intro}

Placeholder


## Plot using base R
## ggplot2
### Histogram
### Bar plot
### Pie chart
### Scatter plot
### arrange plot in grids
### Line plot
### Facet wrap
### Polygons
### Gantt chart
### Heatmap
## ggplot2 extra
### Alluvial and Sankey diagram
### Survival plot
### ggraph and tidygraph
### ggparty-decision tree
### ggROC- ROC curve
### Map 
#### Thematic map
#### tmap
#### Voronoi
### ggwordcloud
### gganimate
### ggneuro
## plotly
### Scatter plot with plotly
### Bar plot with plotly
### Heatmap
### map

<!--chapter:end:01-intro.Rmd-->


# Data Wrangling

Placeholder


## Data 
### Vector, Arrays, Matrix
#### Vector and list
#### Matrix and Arrays
### apply, lapply, sapply
#### apply
#### lapply
#### sapply
#### tapply
### Simple function
### for loop
### Functional
#### Mapply
## Data storage
### Data frame
### Excel data
#### Date and time
### Foreign data
### json format
## Tidy data
### Factors
### Multiple files
### Pivot
## Regular Expressions
### base R
### stringr
## PDF to xcel
### Scanned text or picture
## Web scraping
## Manipulating Medical Images
### DICOM and nifti format
### Manipulating array of medical images
### Combining arrays
### Math operation on multidimensional array
### Math operation on list
### Vectorising nifti object
### tar file
### Image registration
### Rescaling
### MNI template 
#### Atlases
#### AAL atlas
### add text
### add text
#### Eve template
#### Sensorimotor tract atlas

<!--chapter:end:02-Data-Wrangling.Rmd-->


# Statistics

Placeholder


## Univariable analyses
### Parametric tests
### Non-parametric tests
## Regression
### Brief review of matrix
### Linear (least square) regression
### Logistic regression
#### Discrimination and Calibration
#### Measuring Improvement in Regression Models 
#### Shapley value
#### ICE
##### Interaction 
#### Propensity matching
## Special types of regression
### Ordinal regression
### Survival analysis
### Quantile regression
##### Non-negative regression
### Poisson regression
### Conditional logistic regression
### Multinomial modelling
## Sample size estimation
### Proportion
#### Non-inferiority
### Logistic regression
### Survival studies
### Multiple regression
## Interpreting clinical trials
### NNT from ARR
### NNT from odds ratio
### NNT from hazard ratio
### NNT from metaananlysis
## Metaanalysis
### PRISMA
### Conversion of mean and median
### Inconsistency I2
### Metaanalysis of proportion
### Bivariate Metaanalysis
### Metaanalysis of clinical trial.
### Metaregression
#### summary Positive and Negative Likelihood Ratio
## Data simulation

<!--chapter:end:03-Statistics.Rmd-->

# Multivariate Analysis

The following section illustrates the different methods in multivariate analyses. These methods are not to be confused with the more multivariable analyses discussed under Statistics.

## Multivariate regression

Multivariable and multivariate regression are often used interchangeably. Some use the term multivariate when there are more than one dependent variables. Multivariable regression refers to linear, logistic or survival curve analysis in the previous chapter.  Multivariate regression refers to nested models or longitudinal models or more complex type of analyses described below.

### Penalised regression

We used penalised logistic regression (PLR) to assess the relationship between the ASPECTS regions and stroke disability (binary outcome) [@pmid23838753]. PLR can be conceptualized as a modification of logistic regression. In logistic regression, there is no algebraic solution to determine the parameter estimate (β coefficient) and a numerical method (trial and error approach) such as maximum likelihood estimate is used to determine the parameter estimate. In certain situations overfitting of the model may occur with the maximum likelihood method. This situation occurs when there is collinearity (relatedness) of the data. To circumvent this, a bias factor is introduced into the calculation to prevent overfitting of the model. The tuning (regularization) parameter for the bias factor is chosen from the quadratic of the norms of the parameter estimate. This method is known as PLR. This method also allows handling of a large number of interaction terms in the model. We employed a forward and backward stepwise PLR that used all the ASPECTS regions in the analysis, calling on the penalized function in R programming environment. This program automatically assessed the interaction of factors in the regression model in the following manner. The choice of factors to be added/deleted to the stepwise regression was based on the cost complexity statistic. The asymmetric hierarchy principle  was used to determine the choice of interaction of factors. In this case, any factor retained in the model can form interactions with others that are already in the model and those that are not yet in the model. In this analysis, we have specified a maximum of 5 terms to be added to the selection procedure. The significance of the interactions was plotted using a previously described method. We regressed the dichotomized mRS score against ASPECTS regions, demographic variables (such as age and sex), physiological variables (such as blood pressure and serum glucose level) and treatment (rt-PA). The results are expressed as β coefficients rather than as odds ratio for consistency due to the presence of interaction terms.


```{r 04-multivariate-analysis-2, warning=F}

library(mice)

data("BreastCancer",package = "mlbench")
colnames(BreastCancer)

#check for duplicates
sum(duplicated(BreastCancer))

#remove duplicates
#keep Id to avoid creation of new duplicates
#BreastCancer1<-unique(BreastCancer) #reduce 699 to 691 rows

#impute missing data
#m is number of multiple imputation, default is 5
#output is a list
imputed_Data <- mice(BreastCancer, m=5, maxit = 5, method = 'pmm', seed = 500)

#choose among the 5 imputed dataset
completeData <- complete(imputed_Data,2)

#convert multiple columns to numeric
#lapply output a list
BreastCancer2<-lapply(completeData[,-c(11)], as.numeric) #list
BreastCancer2<-as.data.frame(BreastCancer2)
BreastCancer2$Class<-BreastCancer$Class

#convert factor to numeric for calculatin of vif
BreastCancer2$Class<-as.character(BreastCancer2$Class)
BreastCancer2$Class[BreastCancer2$Class=="benign"]<-0
BreastCancer2$Class[BreastCancer2$Class=="malignant"]<-1
BreastCancer2$Class<-as.numeric(BreastCancer2$Class)

BC <- unique(BreastCancer2) # Remove duplicates

#check correlation
library(ggcorrplot)
ggcorrplot(cor(BC),
    p.mat=cor_pmat(BC),hc.order=T, type="lower", colors=c("red","white","blue"),tl.cex = 8)

```

### MARS

Multivariate adaptive regression spline (MARS) is a non-linear regression method that fits a set of splines (hinge functions) to each of the predictor variables i.e. different hinge function for different variables [@pmid8548103]. As such, 
the method can be used to plot the relationship between each variable and 
outcome. Use in this way, the presence of any threshold effect on the predictors can be graphically visualized. The MARS method is implemented in R programming environment in the _earth_ package.

```{r 04-multivariate-analysis-3, warning=F}

library(earth)

BC<-BC[-1]

Fit<-earth(Class ~.,data= BC,
           nfold=10,ncross=30, varmod.method = "none",
           glm=list(family=binomial))

plotmo(Fit)
summary(Fit)
```


### Mixed modelling

In a standard regression analysis, the data is assumed to be random. Mixed 
models assume that there are more than one source of random variability in the data. This is expressed in terms of fixed and random effects. Mixed modeling is 
a useful technique for handling multilevel or group data. The intraclass correlation (ICC) is used to determine if a multilevel analysis is necessary ie 
if the infarct volume varies among the surgeon or not. ICC is the between group variance to the total variance. If the ICC approaches zero then a simple regression model would suffice.

There are several R packages for performing mixed modeling such as _lme4_. Mixed modeling in meta-regression is illustrated in the section on Metaanalysis. An example of mixed model using Bayesian approach with INLA is provided in the [Bayesian section][INLA, Stan and BUGS]

#### Random intercept model

In a random intercept  or fixed slope multilevel model the slope or gradient of the fitted lines are assumed to be parallel to each other and the intercept 
varies for different groups. This can be the case of same treatment effect on animal experiments performed by different technician or same treatment in different clusters of hospitals. There are several approached to performing analysis with random intercept model. The choice of the model depends on the reason for performing the analysis. For example, the maximum likelihood 
estimation (MLE) method is better than restricted maximum likelihood (RMLE) in that it generates estimates for fixed effects and model comparison. RMLE is preferrred if there are outliers. 


#### Random slope model

In a random slope model, the slopes are not paralleled 


### Trajectory modelling

Trajectory analysis attempts to group the behaviour of the subject of interest over time. There are several different approaches to trajectory analysis: data 
in raw form or after orthonal transformation of the data in principal component analysis. Trajectory analysis is different from mixed modelling in that it examines group behaviour. The output of trajectory analysis is only the 
beginning of the modeling analysis. For example, the analysis may identify that there are 3 groups. These groups are labelled as group A, B and C. The next step would be to use the results in a modelling analysis of your choice.

A useful library for performing trajectory analysis is _akmedoids_. This library anchored the analysis around the median value. The analysis requires the data in long format. The _traj_ library is similar to the one in _Stata_. It uses 
several steps including factor and cluster analyses to idetify groups. The 
_traj_ model prefers data in wide format.


### Generalized estimating equation (GEE)

GEE is used for analysis of longitudinal or clustered data. GEE is preferred when the idea is to discover the group effect or population average (marginal) log odds [@pmid20220526]. This is contrast with the mixed model approach to evaluate the average subject via maximum likelihood estimation. The fitting for mixed model is complex compare to GEE and can breakdown. The library for performing GEE is _gee_ or _geepack_.

```{r 04-multivariate-analysis-4, warning=F}
library(tidyverse)
library(gee)

#open simulated data from previous chapter
dtTime<-read.csv("./Data-Use/dtTime_simulated.csv") %>%
  rename(NIHSS=Y)
  

fit<-gee(ENI~T+Diabetes+NIHSS,
         id=id, 
         corstr = "unstructured",
         #data=dtTrial_long)
          data=dtTime)

```

## Principal component analysis

Principal component analysis (PCA) is a data dimension reduction method which 
can be applied to a large dataset to determine the latent variables (principal components) which best represent that set of data. A brief description of the method is described here and a more detailed description of the method can be found in review [@pmid10703049]. The usual approach to PCA involves eigen 
analysis of a covariance matrix or singular value decomposition. 

PCA estimates an orthogonal transformation (variance maximising) to convert a 
set of observations of correlated variables into a set of values of uncorrelated (orthogonal) variables called principal components. The first extracted 
principal component aligns in the direction that contains most of the variance 
of observed variables. The next principal component is orthogonal to the first principle component and contains the second most of spread of variance. The next component contains the third most of spread, and so on.  The latter principal components are likely to represent noise and are discarded. Expressing this in terms of our imaging data, each component yields a linear combination of ‘ischemic’ voxels that covary with each other. These components can be interpreted as patterns of ischemic injury. The unit of measurement in PCA images is the covariance of the data. 

In the case of MR images, each voxel is a variable, leading to tens of thousands of variables with relatively small numbers of samples. Specialised methods are required to compute principal components. 

Based on cosine rule, principal components from different data are similar if values approach 1 and dissimilar if values approach 0 [@pmid22551679].

There are situations in which PCA may not work well if there is non-linear relationship in the data.

```{r PCA}
library(oro.nifti)
library(abind)
library(CHNOSZ) # for working with arrays
library(RNiftyReg)

#create a list using pattern matching
mca.list<-list.files(path="./Ext-Data/",pattern = "*.nii", full.names = TRUE)

#length of list
length(mca.list)
#read multiple files using lapply function

#use lappy to read in the nifti files
#note lapply returns a list
mca.list.nii <- lapply(mca.list, readNIfTI)

#convert to list of array 
mca.list.array<-lapply(mca.list.nii, img_data)

#convert to array
m.listarray<-list2array(mca.list.array)

library(mand)
fit=msma(m.listarray,comp=2)


#https://rdrr.io/cran/mand/f/vignettes/a_overview.Rmd
#https://rdrr.io/cran/caret/man/plsda.html
```


## Independent component analysis

Independent component analysis is different from PCA in that it seeks components which are statistically independent. 

## Partial least squares

There are several versions of partial least squares (PLS). A detailed 
mathematical exposition of the PLS-PLR technique used here can be found in the paper by Fort and Lambert-Lacroix [@pmid15531609]. For the purposes of 
exposition we will describe the individual components of the method. PLS is a multiple regression method that is suited to datasets comprising large sets of independent predictor variables (voxels in an image) and smaller sets of 
dependent variables (neurological outcome scores). Each voxel can take on a 
value of 1 (representing involvement by infarction) or 0 (representing absence 
of involvement) in the MR image of each patient. PLS employs a data reduction method which generates latent variables, linear combinations of independent and dependent variables which explain as much of their covariance as possible. 
Linear least squares regression of the latent variables produces coefﬁcients or beta weights for the latent variables at each voxel location in the brain in stereotaxic coordinate space.[@pmid19660556]

The colon dataset containing microarray data comes with the _plsgenomics_ 
library [@pmid28968879]. The analysis involves partitioning the data into 
training and test set. The classification data is in the Y column. This example 
is provided by the _plsgenomics_ library

```{r 04-multivariate-analysis-5, warning=F}
library(plsgenomics)

data("Colon")
class(Colon) #list
#62 samples 2000 genes
#Outcome is in Y column as 1 and 2. 62 rows
#2000 gene names

dim(Colon$X) 

#heatmap
matrix.heatmap(cbind(Colon$X,Colon$y))

#
IndexLearn <- c(sample(which(Colon$Y==2),12),sample(which(Colon$Y==1),8))
Xtrain <- Colon$X[IndexLearn,] 
Ytrain <- Colon$Y[IndexLearn] 
Xtest <- Colon$X[-IndexLearn,]

# preprocess data 
resP <- preprocess(Xtrain= Xtrain, Xtest=Xtest,Threshold = c(100,16000),Filtering=c(5,500), log10.scale=TRUE,row.stand=TRUE)

# Determine optimum h and lambda
hlam <- gsim.cv(Xtrain=resP$pXtrain,Ytrain=Ytrain,hARange=c(7,20),
              LambdaRange=c(0.1,1),hB=NULL)

# perform prediction by GSIM 
# lambda is the ridge regularization parameter from the cross validation
res <- gsim(Xtrain=resP$pXtrain, 
            Ytrain= Ytrain,Xtest=resP$pXtest,
            Lambda=hlam$Lambda,hA=hlam$hA,hB=NULL)
res$Cvg 

#difference between predicted and observed
sum(res$Ytest!=Colon$Y[-IndexLearn])
```

## Causal inference


<!--chapter:end:04-multivariate-analysis.Rmd-->


# Machine learning

Placeholder


## Decision tree analysis
### Information theory driven
### Conditional decision tree
### criticisms of decision tree
## Random Forest
### Random survival forest
## Gradient Boost Machine
### Extreme gradient boost machine
## KNN
## Support vector machine
### Survival analysis using random forest
## Non-negative matrix factorisation
## Formal concept analysis
## Evolutionary Algorithm
### Simulated Annealing
### Genetic Algorithm
## Manifold learning
### T-Stochastic Neighbourhood Embedding
### Self organising map
## Deep learning
### Multiplayer Perceptron
### CNN
### RNN
### Reinforcement learning

<!--chapter:end:05-machinelearning.Rmd-->


# Machine Learning Part 2

Placeholder


## Bag of words
### TFIDF
### Extracting data from web
## Wordcloud
## Bigram analysis
## Trigram
## Topic modeling or thematic analysis
### Probabilistic topic model
### NMF

<!--chapter:end:06-machinelearningpt2.Rmd-->


# Bayesian Analysis

Placeholder


## Baysian belief
### Conditional probability
#### Bayes Rule
#### Conditional independence
## Markov model
## INLA, Stan and BUGS
### Linear regression
### Logistic regression
### Mixed model
### Bayesian Metaanalysis 
### Cost

<!--chapter:end:07-Bayesian-analysis.Rmd-->


# Operational Research

Placeholder


## Queueing theory
## Discrete Event Simulations
### Simulate capacity of system
### Queuing network
## Linear Programming
## Forecasting
### Bed requirement
### Length of stay
### Customer churns
## Process mapping
## Supply chains
## Health economics
### Cost

<!--chapter:end:08-operational-research.Rmd-->


# Graph Theory

Placeholder


## Special graphs 
### Laplacian matrix
### Bimodal (bipartite) graph
## Centrality Measures
### Local centrality measures
### Global centrality measures
#### Page Rank
## Community
## Visualising graph
### Visnetwork
### Large graph
## Social Media and Network Analysis
### Twitter
### Youtube

<!--chapter:end:09-graph-theory.Rmd-->


# Geospatial analysis

Placeholder


## Geocoding
### OpenStreetMap
### Google Maps API
## Sp and sf objects
## Thematic map
### Calculate distance to Hospital-OpenStreetMap
## Spatial regression
### New York COVID-19 mortality
### Danish Stroke Registry
### INLA
### Stan
## Machine learning in spatial analysis
## Spatio-temporal regression

<!--chapter:end:10-geospatial-analysis.Rmd-->


# App

Placeholder


## Brief introduction to Shiny app

<!--chapter:end:11-App.Rmd-->


# Appendix

Placeholder


## Brief introduction to Matrix
## Regression
### Linear regression
#### Leasts squares
#### Collinearity
#### weighted least squares
### Logistic regression

<!--chapter:end:12-Appendix.Rmd-->

`r if (knitr:::is_html_output()) '
# References {-}
'`

<!--chapter:end:13-references.Rmd-->

