--- 
title: "Applications of R in Healthcare Book"
author: "Thanh Phan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a book describing the use of R in Healthcare. It is aimed at beginners of R language. The majority of the examples are taken from works in Neurology. The ideas and principles can be applied to other aspects of Healthcare. The output format for this example is bookdown::gitbook."
---

# Preamble

This is a _book_ adapted from Yihui Xie's package bookdown written in **Markdown**. R can be installed from The Comprehensive R Archive Network at https://cran.r-project.org/. It is _free_. The user can choose between three different platforms such as Windows, Mac (OS X) and Linux.Following installation of R, you should install Rstudio, available at https://rstudio.com/.

```{r eval=FALSE}
#write in bold by using **
**Markdown**

#write in italics by using _
_book_

#equations can be done using $
#this equation typifies Pythagoras' theorem.
#It has use in least squares approach to regression
$Z^2=x^2+y^2$
```

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

